---
output: 
  revealjs::revealjs_presentation:
    transition: "none"
    font_adjustment: +20
---

## Main components of an R-package

<br>
```{r, out.width = "500pt", echo = FALSE, fig.align="center"}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/Rlogo.jpg")
```



## Why R is so useful
<style>
    .reveal .slides {
        text-align: left;
    }
    .reveal .slides section>* {
        margin-left: 0;
        margin-right: 0;
    }
    .reveal .slides section .slideContent{
        font-size: 62px;
    }
    .reveal section img {
    border: none;
    box-shadow: none;
  }
</style>

<style type="text/css">
{
  font-size: 28px;
}
</style>


<br>
<br>
Easy to use

<br>
Easy to get help

<br>
A lot of work has been done already

## Why R is so useful

<br>
<br>
Easy to use

<br>
Easy to get help

<br>
**A lot of work has been done already**


## What is an R package (for the user)

<br>
A set of functions or data!
<br>
```{r}
library(raster)

raster()
```

## Famous packages

```{r, out.width = "1000in", echo = FALSE}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/Popular-R-Packages-2.png")
```


## Famous packages in ecology

<br>
~15,000 packges on CRAN!

<br>
vegan

lme, INLA

raster, sp

ape, phytools

## The purpose of R packages

<br>
Distribute code

Distribute data

Organize analyses

Collaborate on a project


## Where packages live

<br>
1. The Comprehensive R Archive Network (CRAN)

<br>
2. GitHub

## 1. The Comprehensive R Archive Network (CRAN)

```{r, out.width = "1000in", echo = FALSE}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/cran_full.png")
```

## 1. The Comprehensive R Archive Network (CRAN)

```{r, out.width = "1000in", echo = FALSE}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/CRAN_packages.png")
```


## We have all done it

<br>
```{r, eval = FALSE}
install.packages("ggplot2")
```

## 2. GitHub
```{r, out.width = "1000in", echo = FALSE}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/github_intro.png")
```

## 2. GitHub
```{r, out.width = "1000in", echo = FALSE}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/github_package.png")
```

## Also easy
<br>
```{r, eval = FALSE}
library(devtools)
install_github("azizka/bromeliad")
```

## Developing an R package

## An example from from the Bromeliaceae

```{r, out.width = "1000in", echo = FALSE}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/pineapple2.png")
```

## An example from from the Bromeliaceae
<br>
```{r, out.width = "1000in", echo = FALSE}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/bromeliad_manuscript.png")
```

## An example from from the Bromeliaceae

```{r, message = FALSE, warnings = FALSE}
library(bromeliad)

data("taxonomy_traits")

taxonomy_traits
```

## An example from from the Bromeliaceae

```{r, message = FALSE, warnings = FALSE}
data("all_ranges")

all_ranges
```

## An example from from the Bromeliaceae
```{r, message = FALSE, warnings = FALSE}
aechmea <- get_range(genus = "Aechmea")
```

<!-- ## An example from from the Bromeliaceae -->
<!-- ```{r, message = FALSE, warnings = FALSE} -->
<!-- map_richness(aechmea) -->
<!-- ``` -->

## An example from from the Bromeliaceae

```{r, message = FALSE, warnings = FALSE}
crvu <- get_range(assessment = c("CR", "VU"))
#map_richness(crvu)
```


## What is an R package (for a developer)
<br>
A folder...

...with standardized structure and names


## Package structure
<br>
`DESCRIPTION`

`README.md`

`NAMSPACE`

`R/`

`man/`

`data/`

<font color="grey">`inst/, vignettes/, tests/ .Rbuildignore,`</font> 
<font color="grey">`.gitignore, cran-comments.md, NEWS.md`</font>


## Package structure

```{r, out.width = "1000pt", echo = FALSE}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/package_folder.png")
```


## DESCRIPTION

A .txt file giving basic and standardized information on the package

## DESCRIPTION
<br>
```
Package: bromeliad
Title: Distribution Data for 3272 Bromeliad Species
Version: 0.9-2
Authors@R: person("Alexander", "zizka", 
                  email = "zizka.alexander@gmail.com",
                  role = c("aut", "cre"))
Description: Contains range estimates and modelled distributions for 
             3272 species of the ananas family (Bromeliaceae), 
             together with some convenience functoins to extract 
             specific taxa and life forms and visualize diversity.
Depends: R (>= 3.5.0)
Imports: dplyr, ggplot2, ggthemes, magrittr, methods,
         raster, rnaturalearth, 
         speciesgeocodeR, rlang, sf, sp, viridis
License: GPL-3
Encoding: UTF-8
RoxygenNote: 6.1.1
```

## README.md
<br>
A description of the package for Github


## README.md
```[md}
# bromeliad 0.9-2

A data package with distribution data for 2272 bromeliaceae species across tropical America.

## Included data
The package includes five data files:

1. `taxonomy traits` - a data.frame with taxonomy, traits and conservation assessment for the included species.
2. `all_ranges` - a simple features object containing range polygons for all 2395 species.
3. `model_binary` - a list of presence/absence rasters, for 541 species for which the ranges were estimated based on climate suitability models
4. `model_ensemble` - a list of rasters with habitat suitability, for 541 species for which the ranges were estimated based on climate suitability models
5. `digitized_records` - a data.frame with the point locations of XXX species type specimens, georeferenced from the literature for this package

## Installation
The package can be installed from github using `devtools::install_github`
library(devtools)

install_github("idiv-biodiversity/bromeliad")

library(bromeliad)


## Documentation
The package also includes some convinieince functions to select taxa and visualize species ranges and diversity. A vignette describing how to select individual taxa and visualize species distributions and diversity patterns is available [here]().

## Contribute
If you want to contribute with new species or corrections, please contact the [maintainer](mailto:alexander.zizka@idiv.de) of the package.

## Licence
The package and data are available under a GPL-3 license.

## Citation
Please cite the package as Zizka A *et al.* (2019) bromeliad. An R package for biogeography and conservations tatus of Bromeliaceae.

```

## README.md
```{r,out.width = "1000pt", echo = FALSE}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/readme_screenshot.png")
```


## NAMESPACE
<br>
A file with the names of all functions that your package uses and provides. 

## NAMESPACE

```
export(get_range)
export(map_richness)
export(map_species)
importFrom(dplyr,filter)
importFrom(dplyr,mutate)
importFrom(ggplot2,aes)
importFrom(ggplot2,annotate)
importFrom(ggplot2,coord_fixed)
importFrom(ggplot2,element_line)
importFrom(ggplot2,fortify)
importFrom(ggplot2,geom_polygon)
importFrom(ggplot2,geom_sf)
importFrom(ggplot2,geom_tile)
importFrom(ggplot2,ggplot)
importFrom(ggplot2,ggtitle)
importFrom(ggplot2,theme)
importFrom(ggplot2,xlim)
importFrom(ggplot2,ylim)
importFrom(ggthemes,theme_map)
importFrom(magrittr,"%>%")
importFrom(methods,as)
importFrom(raster,crop)
importFrom(raster,extend)
importFrom(raster,extent)
importFrom(raster,raster)
importFrom(raster,rasterToPoints)
importFrom(rlang,.data)
importFrom(rnaturalearth,ne_download)
importFrom(sf,st_transform)
importFrom(sp,CRS)
importFrom(sp,spTransform)
importFrom(speciesgeocodeR,RangeRichness)
importFrom(utils,data)
importFrom(viridis,scale_fill_viridis)
```

## data/
<br>
The datasets distributed with the package. Can be example data or the purpose of the package.

## data/
<br>
Can be any object that can be loaded into R.

Tables, shape files, ...

## data/

```{r, message = FALSE, warnings = FALSE}
data("taxonomy_traits")

taxonomy_traits
```

## R/
<br>
The code (functions) provided via your package.

## Functions
<br>
In packages code is distributed as functions.
<br>
Functions are objects just as variables.

## Functions
```{r, eval = F}
function_name(x, options)
```

```{r}
x <- c(2,3,5,6)

mean(x)
```

## For example
```{r}
sum_numbers <- function(x, y){
  out <- x + y
  return(out)
 }

sum_numbers(x = 2, y = 2)

sum_numbers(x = 1000, y = 1)
```


## For example
```{r}
smiley <- function(happy){
  if(happy){
    print(":-)")
  }else{
    print(":-(")
  }
}

smiley(happy = TRUE)

smiley(happy = FALSE)
```

## R/
```{r, out.width = "1000pt", echo = FALSE}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/R_folder.png")
```


## R/

```
get_range <- function(scientific = NULL, canonical = NULL,
                      genus = NULL, subfamily = NULL,
                      life_form = NULL,
                      assessment = NULL, range_size = NULL,
                      type = "polygon"){

  match.arg(type, choices = c("binary", "suitability", "polygon"))

  taxonomy_traits <- bromeliad::taxonomy_traits

  if(!is.null(scientific)){
    taxonomy_traits <- dplyr::filter(taxonomy_traits, .data$tax_accepted_name %in% scientific)
  }

  if(!is.null(canonical)){
    taxonomy_traits <- dplyr::filter(taxonomy_traits, .data$tax_canonical %in% canonical)
  }

  if(!is.null(genus)){
    taxonomy_traits <- dplyr::filter(taxonomy_traits, .data$tax_genus %in% genus)
  }

  if(!is.null(subfamily)){
    taxonomy_traits <- dplyr::filter(taxonomy_traits, .data$tax_subfamily %in% subfamily)
  }

  if(!is.null(life_form)){
    taxonomy_traits <- dplyr::filter(taxonomy_traits, .data$tr_growth_form %in% life_form)
  }

  if(!is.null(assessment)){
    taxonomy_traits <- dplyr::filter(taxonomy_traits, .data$conv_conr_stat %in% assessment)
  }

  if(!is.null(range_size)){
    taxonomy_traits <- dplyr::filter(taxonomy_traits, .data$geo_range > range_size[1] & .data$geo_range < range_size[2])
  }

  switch(type,
         binary = {model_binary <-  bromeliad::model_binary;
                   out <- model_binary[which(names(model_binary) %in% taxonomy_traits$tax_accepted_name)]},
         suitability = {model_ensemble <- bromeliad::model_ensemble; out <- model_ensemble[which(names(model_ensemble) %in% taxonomy_traits$tax_accepted_name)]},
         polygon = {all_ranges <- bromeliad::all_ranges; out <- dplyr::filter(all_ranges, .data$tax_accepted_name %in% taxonomy_traits$tax_accepted_name)}
           )

  class(out) <- c("bromeli", class(out))

  return(out)
}


```

## man/
<br>
The documentation of your code. One file per function.

## man/
```{r, out.width = "1000pt", echo = TRUE, eval = F}
?get_range
```

```{r, out.width = "800pt", fig.align="center", echo = FALSE}
knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/function_help.png")
```

## man/
```{r, out.width = "1000pt", echo = FALSE}

knitr::include_graphics(path = "C:/Users/az64mycy/Dropbox (iDiv)/teaching/2020_yDiv_Rpackaging/yDiv_building_packages_Rstudio_GitHub/lectures/pictures/man_folder.png")
```


## man/
```
#' Select Bromeliaceae Species Ranges by Taxonomy and Traits
#'
#'Get the geographic range for all species selected via the arguments.
#'The type of range estimate depends on the \dQuote{type} argument.
#'
#'Modelled ranges are available for 542 species,
#'range polygons for 2395 species. For species with model distribution, the range polygons are based on the models, otherwise
#'on a convex hull around available occurrence records, or a 50 km buffer for species with less than 3 occurrence records available.
#'See Zizka et al 2019 for methods.
#'
#' @param scientific a character vector of full scientific names including authorities, of the species of interest
#' @param canonical a character vector of canonical names, of the species of interest.
#' @param genus a character vector of genera names to select.
#' @param subfamily a character vector of subfamily names to select.
#' @param life_form a character vector of life forms to select.
#' @param assessment a character vector of conservation assessment to select=.
#' @param range_size a vector of two numericals with the minimum and maximum range size (in square kilometres), to select.
#' @param type a cahracter defining the output format, see details
#'
#' @return Depending on the \dQuote{type} argument. If \dQuote{binary} a presence/absence raster based on the modelled habitat suitability,
#'  at 100x100 km resolution in Behrmann projection,
#'  if \dQuote{suitability} the habitat suitability as predicted by an ensemble model in Behrmann projection, and
#'  if {polygon} a simple feature object in lat/lon projection.
#' .
#' @examples
#' get_range(scientific = "Aechmea mexicana Baker")
#' get_range(scientific = "Aechmea mexicana Baker", type = "binary")
#' get_range(scientific = "Aechmea mexicana Baker", type = "suitability")
#' get_range(canonical = "Aechmea mexicana")
#' get_range(genus = "Aechmea")
#' get_range(genus = "Aechmea", type = "binary")
#' get_range(genus = c("Aechmea", "Zizkaea"))
#' get_range(subfamily = "Pitcairnioideae")
#' get_range(life_form = "epiphyte")
#' get_range(assessment = c("CR", "VU"))
#' get_range(range_size = c(1000, 10000))
#'
#' @export
#' @importFrom dplyr filter
#' @importFrom methods as
#' @importFrom rlang .data
#'
#'
```

## The rest

<br>
`inst/` - non-text files (e.g. pdfs)

`vignettes/` - Tutorials and documentation

`test` - automated tests of the code

`.gitignore` - files git should ignore

`.Rbuildignore` - files R CMD build should ignore


## Build and check

<br>
Build

Check

Deploy


## Summary

<br>
An R package is a set of standardized folders

`Description` - basic information

`Readme` - basic information

`data/` -  datasets of the package

`R/` - the functions of the package

`man/` - the documentation of the code

others - to facilitate the developement progress and provide additional information

## How to develope an R-package
<br>
<br>
Rstudio, Github, Git, devtools, roxygen2

<br>
Rstudio workflow. Next lecture!

## Your project!
<br>
<br>
Chose a dataset from your research that you want to provide as an R package, and one or two basic functions to use it.

## Remember
<br>
Error messages are you frineds!


