---
title: "Package testing"
author: 
date: "September 24, 2019"
output: 
  revealjs::revealjs_presentation:
    transition: "none"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

report_summary <- function(data, measure, na.rm = TRUE, quantiles = NULL){
  
  if(measure == "mean"){
    out <- mean(x = data, na.rm = na.rm)
  } else if(measure == "median"){
    out <- median(x = data, na.rm = na.rm)  
  }
  out <- round(x = out, digits = 2)
  
  if(!is.null(quantiles)){
    out <- c(out, quantile(data, probs = quantiles, na.rm = na.rm))
  }
  names(out)[1] <- measure
  
  return(out)
}
library(checkmate)
```

## What is "testing"

1. checks
2. assertions
3. tests

## 1. checks

`check()` <br>

run a range of diagnostic checks to ensure validity of the package (setup, metadata)

make sure that examples can be run

however ...

## Invalid arguments?

```{r}
greet <- function(whom){
  out <- paste0("Hello '", whom, "'!")
  return(out)
}
aList <- list(people = c("Amelie", "Peter"), animals = c("my cat"))
```
<div class="fragment">

```{r}
greet(whom = aList)
```
</div>

## Invalid arguments!

```{r, error=TRUE}
report_summary(data = iris, measure = "median")
report_summary(data = iris$Sepal.Length, measure = "sd")
```

## 2. assertions

`checkmate::assert*()` <br>

make sure that arguments fullfil a set of expectations, otherwise throw an error.

standardised interface, where specific argument types require specific options.


## 2. assertions

`assertVector` -> Check if an argument is a vector

<div class="fragment">
```{r, error=TRUE}
data <- iris
assertVector(x = data)

assertVector(x = data, strict = TRUE)
```
</div>

## 2. assertions

`assertChoice` -> Check if an object is an element of a given set

```{r, error=TRUE}
measure <- NULL
assertChoice(x = measure, choices = c("mean", "median"), null.ok = TRUE)

measure <- "sd"
assertChoice(x = measure, choices = c("mean", "median"))
```

## 2. assertions

`assertLogical` -> Check if an argument is a vector of type logical

```{r, error=TRUE}
na.rm <- NA
assertLogical(x = na.rm)

assertLogical(x = na.rm, any.missing = FALSE)
```
## 2. assertions

`assertNumeric` -> Check that an argument is a vector of type numeric

```{r, error=TRUE}
quantiles <- c(0, 0.5, NA, 2)
assertNumeric(x = quantiles, lower = 0, upper = 1, any.missing = FALSE)
```
<div class="fragment">
```{r, error=TRUE}
quantiles <- c(0, 0.5, 2)
assertNumeric(x = quantiles, lower = 0, upper = 1, any.missing = FALSE)
```
</div>


## 2. assertions
```{r, include=FALSE}
report_summary <- function(data, measure, na.rm = TRUE, quantiles = NULL){
  
  assertVector(x = data, strict = TRUE)
  assertChoice(x = measure, choices = c("mean", "median"))
  assertLogical(x = na.rm, any.missing = FALSE, len = 1)
  assertNumeric(x = quantiles, lower = 0, upper = 1, any.missing = FALSE, null.ok = TRUE)
  
  if(measure == "mean"){
    out <- mean(x = data, na.rm = na.rm)
  } else if(measure == "median"){
    out <- median(x = data, na.rm = na.rm)  
  }
  out <- round(x = out, digits = 2)
  
  if(!is.null(quantiles)){
    out <- c(out, quantile(data, probs = quantiles, na.rm = na.rm))
  }
  names(out)[1] <- c(measure)
  
  return(out)
}
```

```{r, error=TRUE}
report_summary(data = iris, measure = "median")
report_summary(data = iris$Sepal.Length, measure = "sd")
```

## 3. tests

initiate testiing with `use_testthat()` <br>

make a test with `use_test()`

set up expectations for your functions and compare them against the status quo.

## 3. tests

```{r, eval=FALSE}
use_test("greet")
```

this opens a new file, in which we find a first test

```{r, eval=FALSE}
test_that("multiplication works", {
  expect_equal(object = 2 * 2, expected = 4)
})
```

<div class="fragment">
change that to

```{r, eval=FALSE}
test_that("output is a character string", {
  out <- greet(whom = "World")
  expect_character(x = out, len = 1, pattern = "Hello.*!")
})
```
</div>

## 3. tests

Finally, run `test()` in the console and get information on how your tests are performing!
