---
title: "Exercise 1 - Package skeleton"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

During the course we will use the RStudio Graphical User Interface. First, we will set up the RStudio project and the connection between RStudio, Git, and GitHub.

# 1. Create a GitHub repository

Go to www.github.com and log in with your account. Create a new repository, with a suitable repository name (ideally the name of your package, only lower-case letters).

# 2. Link an Rstudio project with GitHub

Open Rstudio and create a new project. Select "Version control" > Git and copy the url of your GitHub repository. Let Rstudio restart. You will now see the git tab newly added to the project panes.

# 3. Set up package build tools.

Go to "Tools" > "Project options" > "Build tools" select "package". Also, make sure to tick all options under `Generate documentation with Roxygen`.

# 4. Set up package structure

As introduced in the lecture, R packages (as long as no compiled code is involved) are a combination of text files and folders in a predefined structure. From now on we will exclusively use the R console to create the files and folders of this structure and step-after-step build the R package. Each step of the building process will be based on an R function that carries out the necessary actions. For many steps there are alternative ways, including to click on some button, but using the R console is the more reproducible and the learning outcome is in our eyes higher. Furthermore, GUI-Buttons may be at different locations in the specific IDE (Integrated development environment, here RStudio) but the function works in any instance of R that has the respective package installed. To provide fully transparent and reproducible documentation, you can write down all those steps you took to make the package via R functions, in a markdown document just like this.

Nowadays, R-package development relies heavily on two R-package that provide the necessary code: the `devtools` and the `usethis` package. To start building a package, we thus load them.


```{r}
library(devtools)
```

## Create description
The description file contains all basic information about your package, and is of crucial importance for the use and distribution of your package. Some say, it is the most important part of the package! The `use_description` function creates template file which you will need to fill.

```{r, eval = FALSE}
usethis::use_description()
```

# Edit `DESCRIPTION`

The `DESCRIPTION` gather all information that crucially important as metadata to your package. It contains, amongst others, title and description, dependencies, the package version and author information. The version created by `create_package()` is relatively self-explanatory.

```{r, eval=FALSE}
Package: myPackage
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: 
    person(given = "First",
           family = "Last",
           role = c("aut", "cre"),
           email = "first.last@example.com",
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: What the package does (one paragraph).
License: GPL-3
Encoding: UTF-8
LazyData: true
RoxygenNote: 6.1.1
```

You have to add some fields yourself, for example the url of the package and where bugs can be reported

```{r eval=FALSE}
URL: https://github.com/"YOUR USERNAME/myPackage"
BugReports: https://github.com/"YOUR USERNAME/myPackage"/issues
```


## Create namespace file
The namespsce is another crucial part of your package. Simply put: the namespace give information on the names used in your pacakge, taht is the functions, datasets, methods etc it exports (introduces) and imports from other packages. We will set up a sceleton namespace with `use_namespace`.

```{r, eval = FALSE}
usethis::use_namespace()
```

## Set a license

Chose the license you want to use for this package and supply it with one of the `use_*_license()` functions.

```{r, eval=FALSE}
usethis::use_gpl3_license()
# ✔ Adding 'GPL (>= 3)' to License
# ✔ Writing 'LICENSE.md'
# ✔ Adding '^LICENSE\\.md$' to '.Rbuildignore'
```


## Dependencies
Some fields will be added automatically with other `usethis` functions. For example, most packages will require functions from other packages that are not part of base R. Let's say you want to do some data wrangling with `dplyr`, want to use `tibble` instead of `data.frame` and intend to describe the idea of your package in a vignette names `overview`. You would include those via.

```{r, eval=FALSE}
usethis::use_package("dplyr")
# ✔ Adding 'dplyr' to Imports field in DESCRIPTION
# • Refer to functions with `dplyr::fun()`
usethis::use_package("tibble")
# ✔ Adding 'tibble' to Imports field in DESCRIPTION
# • Refer to functions with `tibble::fun()`
usethis::use_vignette(name = "overview")
# ✔ Adding 'knitr' to Suggests field in DESCRIPTION
# ✔ Adding 'rmarkdown' to Suggests field in DESCRIPTION
# ✔ Adding 'knitr' to VignetteBuilder
# ✔ Adding 'inst/doc' to '.gitignore'
# ✔ Creating 'vignettes/'
# ✔ Adding '*.html', '*.R' to 'vignettes/.gitignore'
# ✔ Writing 'vignettes/overview.Rmd'
# • Modify 'vignettes/overview.Rmd'
```

This would add the following to `DESCRIPTION`, right after `RoxygenNote: 7.2.3`.

```{r eval=FALSE}
Imports: 
    dplyr,
    tibble
Suggests: 
    knitr,
    rmarkdown
VignetteBuilder: knitr
```

The packages listed in `Imports:` have to be installed on the users system and the packages listed in `Suggests:` are not required, but optionally used at some point in your package. `knitr` and `rmarkdonw` will, for example, be used when you build the vignette, but the user does not have to do that on their machine.

## Versioning

`use_description()` has already assigned a version to your package. Whenever you include additions, you should increase that version number. For now it is sufficient that each new feature increases the number in your `DESCRIPTION` by one, for example from `0.0.0.9000` to `0.0.0.9001`. For more detailed information on how to handle versioning, check out [this](https://r-pkgs.org/description.html#version).

## Include your ORCiD

`create_package()` has already included a field in `DESCRIPTION` for your ORCiD identifier. This is recently a comment to your name. Make use of it for all authors!



<!-- # Create the package -->

<!-- You can start setting up a project via clicking on *File > New Project > New Directory > R package* and fill in information in the dialog that opens. However, you can also simply use the function `create_package()`, provide it with the path where you want to package to be stored and let it set up everything. Realistically, using the command line may even be faster when you are not used to it (yet). -->

<!-- ```{r, eval=FALSE} -->
<!-- create_package(path = "/path/to/myPackage") -->
<!-- #> ✔ Creating '/path/to/myPackage/' -->
<!-- #> ✔ Setting active project to '/path/to/myPackage' -->
<!-- #> ✔ Creating 'R/' -->
<!-- #> ✔ Writing 'DESCRIPTION' -->
<!-- #> Package: myPackage -->
<!-- #> Title: What the Package Does (One Line, Title Case) -->
<!-- #> Version: 0.0.0.9000 -->
<!-- #> Authors@R (parsed): -->
<!-- #>     * First Last <first.last@example.com> [aut, cre] (<https://orcid.org/YOUR-ORCID-ID>) -->
<!-- #> Description: What the package does (one paragraph). -->
<!-- #> License: What license it uses -->
<!-- #> Encoding: UTF-8 -->
<!-- #> LazyData: true -->
<!-- #> ✔ Writing 'NAMESPACE' -->
<!-- #> ✔ Writing 'myPackage.Rproj' -->
<!-- #> ✔ Adding '.Rproj.user' to '.gitignore' -->
<!-- #> ✔ Adding '^myPackage\\.Rproj$', '^\\.Rproj\\.user$' to '.Rbuildignore' -->
<!-- #> ✔ Opening '/path/to/myPackage/' in new RStudio session -->
<!-- #> ✔ Setting active project to '/path/to/teaching/yDiv_building_r_packages' -->
<!-- ``` -->

<!-- The grey text is printed to the console to inform you about the progress. After it has been successful, a new RStudio session will be started, in which you would commence package development. This is it already, the most basic package skeleton has been created at `/path/to/myPackage/`. -->


<!-- # Set up git(hub) -->

<!-- As a next step we already set up git, to trace the development of our package and to be able to revert to a previous state if we have included a grave mistake that can't be corrected easily. -->

<!-- ```{r, eval=FALSE} -->
<!-- library(devtools) -->
<!-- use_git() -->
<!-- #> ✔ Setting active project to '/path/to/myPackage' -->
<!-- #> ✔ Initialising Git repo -->
<!-- #> ✔ Adding '.Rhistory', '.RData' to '.gitignore' -->
<!-- #> There are 5 uncommitted files: -->
<!-- #> * '.gitignore' -->
<!-- #> * '.Rbuildignore' -->
<!-- #> * 'DESCRIPTION' -->
<!-- #> * 'myPackage.Rproj' -->
<!-- #> * 'NAMESPACE' -->
<!-- #> Is it ok to commit them? -->
<!-- #>  -->
<!-- #> 1: Negative -->
<!-- #> 2: For sure -->
<!-- #> 3: No way -->
<!-- #>  -->
<!-- #> Selection: 2 -->
<!-- #> ✔ Adding files -->
<!-- #> ✔ Commit with message 'Initial commit' -->
<!-- #> ● A restart of RStudio is required to activate the Git pane -->
<!-- #> Restart now? -->
<!-- #>  -->
<!-- #> 1: Nope -->
<!-- #> 2: I agree -->
<!-- #> 3: Absolutely not -->
<!-- #>  -->
<!-- #> Selection: 2 -->
<!-- ``` -->

<!-- Which would restart RStudio once again to update its interface for git-functionality (i.e., add buttons to access and manage git). Call `library(devtools)` once again. Configure your `git` data by entering the following information into the Terminal, not the Console, into which you have entered the R code so far. The Terminal is in RStudio right next to the Console Tab. The Email address must be the same as the one you used to register your github account. -->

<!-- ```{r, eval=FALSE} -->
<!-- git config --global user.name "YOUR FULL NAME" -->
<!-- git config --global user.email "YOUR EMAIL ADDRESS" -->
<!-- ``` -->

<!-- Now set up a github repository (at [www.github.com](www.github.com)), into which the code will be pushed and tell R about it via the following code (again in the Terminal). -->

<!-- ```{r, eval=FALSE} -->
<!-- git remote add origin git@github.com:"YOUR USERNAME/myPackage".git -->
<!-- git push -u origin master -->
<!-- ``` -->


